---
title: "Ligand-receptor signalling inference"
author: Jamie Soul
date: today
date-format: short
format:
  html:
    self-contained: true
    theme: litera
    toc: true
editor: visual
code-block-bg: true
code-block-border-left: "#31BAE9"
---

# Ligand-receptor signalling inference

This notebook uses NicheNet to infer the t-cell ligand to keratinocyte receptor signalling driving the observed differential expression.

## Load libraries

```{r}
#| output: false
library(NanoStringNCTools)
library(nichenetr)
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(ggpubr)
library(writexl)
library(ComplexHeatmap)
library(patchwork)

source("src/utilityFunctions.R")
set.seed(123)
options(timeout = 100000000000)
```

## Download the ligand-target matrix

```{r}
if(!file.exists("data/ligand_target_matrix_nsga2r_final.rds")){
ligand_target_matrix <- readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds?download=1"))
} else {
  ligand_target_matrix <- readRDS("data/ligand_target_matrix_nsga2r_final.rds")
}


ligand_target_matrix[1:5,1:5]
```

## Define the expressed ligands and receptors

```{r}
#load the normalised expressed data
target_spatialData <- readRDS("results/normalisedSpatialData.RDS")

#determine the expressed genes in atopic dermatitis t-cells
expressed_genes_sender_AD <- target_spatialData[ ,pData(target_spatialData)$Annotation %in% c("CD8_AD","CD4_AD")] %>%
  assayDataApply(2, FUN = log, base = 2, elt = "q_norm") %>%
  which(x = rowMeans(.)>1) %>%
  names()

#determine the expressed genes in atopic dermatitis keratinocytes
expressed_genes_receiver_KRT_AD <- target_spatialData[ ,pData(target_spatialData)$Annotation=="KRT_AD"] %>%
  assayDataApply(2, FUN = log, base = 2, elt = "q_norm") %>% 
  which(x = rowMeans(.)>1) %>% 
  names()

#determine the expressed genes in psoriasis t-cells
expressed_genes_sender_P <- target_spatialData[ ,pData(target_spatialData)$Annotation %in% c("CD8_P","CD4_P")] %>%
  assayDataApply(2, FUN = log, base = 2, elt = "q_norm") %>%
  which(x = rowMeans(.)>1) %>%
  names()

#determine the expressed genes in psoriasis keratinocytes
expressed_genes_receiver_KRT_P <- target_spatialData[ ,pData(target_spatialData)$Annotation=="KRT_P"] %>%
  assayDataApply(2, FUN = log, base = 2, elt = "q_norm") %>%
  which(x = rowMeans(.)>1) %>%
  names()
  
```

## Define the differentially expressed genes

```{r}

geneset_AD = readRDS("results/diffExpLME4_Q3.RDS") %>% 
  filter(Contrast=="KRT_AD - KRT_H") %>%
  filter(FDR<=0.05 & abs(Estimate)>=log2(1.5)) %>% 
  pull(Gene)

geneset_P = readRDS("results/diffExpLME4_Q3.RDS") %>% 
  filter(Contrast=="KRT_P - KRT_H") %>%
  filter(FDR<=0.05 & abs(Estimate)>=log2(1.5)) %>%
  pull(Gene)
```

## Run NicheNet

```{r}
nicheNet_AD <- runNicheNetAnalysis(expressed_genes_sender_AD,expressed_genes_receiver_KRT_AD,geneset_AD,senderName = "Prioritized CD8+ ligands",receiverName = "KRT_AD")

nicheNet_P <- runNicheNetAnalysis(expressed_genes_sender_P,expressed_genes_receiver_KRT_P,geneset_P,senderName = "Prioritized CD8+ ligands",receiverName = "KRT_P")

```

## Plot results

### Atopic dermatitis

The histogram of activity scores and the threshold for the atopic dermatitis is shown.

```{r}
#| label: fig-ligandActivityHistogramAD
#| fig-cap: Histogram of inferred ligand activity in atopic dermatitis
#| fig-width: 8
#| fig-height: 8
#| warning: false
nicheNet_AD$ligandActivityTable %>% arrange(desc(aupr_corrected)) %>%
  write_xlsx( path = "results/ligandActivity_AD.xlsx", col_names = TRUE)

save_plot("figures/LigandToReceptor/ligandHistogramAD.png",nicheNet_AD$ligandActivityPlot,base_height = 5,base_width = 5,bg="white")

nicheNet_AD$ligandActivityPlot

```


```{r}
#| label: fig-ligandActivityTopAD
#| fig-cap: Top inferred ligand activity in atopic dermatitis
#| fig-width: 8
#| fig-height: 8
#| warning: false

save_plot("figures/LigandToReceptor/topLigandsAD.png",nicheNet_AD$ligandPearson,base_height = 5,base_width = 6,bg="white")

nicheNet_AD$ligandPearson
```

```{r}
#| label: fig-ligandTargetsAD
#| fig-cap: Histogram of top inferred ligand to targets potentials in atopic dermatitis
#| fig-width: 8
#| fig-height: 8
#| warning: false

save_plot("figures/LigandToReceptor/LigandsToTargetsAD.png",nicheNet_AD$ligandTarget,base_height = 5,base_width = 5,bg="white")
nicheNet_AD$ligandTarget
```

```{r}
#| label: fig-ligandReceptorAD
#| fig-cap: Histogram of top inferred ligand to targets potentials in atopic dermatitis
#| fig-width: 8
#| fig-height: 8
#| warning: false

save_plot("figures/LigandToReceptor/LigandsToReceptorAD.png",nicheNet_AD$ligandReceptor,base_height = 5,base_width = 5,bg="white")
nicheNet_AD$ligandReceptor

```

### Psoriasis

The histogram of activity scores and the threshold for the psoriasis is shown.

```{r}
#| label: fig-ligandActivityHistogramP
#| fig-cap: Histogram of inferred ligand activity in atopic dermatitis
#| fig-width: 8
#| fig-height: 8
#| warning: false
#
nicheNet_P$ligandActivityTable %>% arrange(desc(aupr_corrected)) %>%
write_xlsx(path = "results/ligandActivity_Psoriasis.xlsx", col_names = TRUE)


save_plot("figures/LigandToReceptor/ligandHistogramP.png",nicheNet_P$ligandActivityPlot,base_height = 5,base_width = 5,bg="white")

nicheNet_P$ligandActivity

```

```{r}
#| label: fig-ligandActivityTopP
#| fig-cap: Top inferred ligand activity in atopic dermatitis
#| fig-width: 8
#| fig-height: 8
#| warning: false

save_plot("figures/LigandToReceptor/topLigandsP.png",nicheNet_P$ligandPearson,base_height = 5,base_width = 6,bg="white")

nicheNet_P$ligandPearson
```

```{r}
#| label: fig-ligandTargetsP
#| fig-cap: Histogram of top inferred ligand to targets potentials in atopic dermatitis
#| fig-width: 8
#| fig-height: 8
#| warning: false

save_plot("figures/LigandToReceptor/LigandsToTargetsP.png",nicheNet_P$ligandTarget,base_height = 5,base_width = 5,bg="white")
nicheNet_P$ligandTarget
```

```{r}
#| label: fig-ligandReceptorP
#| fig-cap: Histogram of top inferred ligand to targets potentials in atopic dermatitis
#| fig-width: 8
#| fig-height: 8
#| warning: false

save_plot("figures/LigandToReceptor/LigandsToReceptorP.png",nicheNet_P$ligandReceptor,base_height = 5,base_width = 5,bg="white")
nicheNet_P$ligandReceptor

```

### Combined results

```{r}

combinedActivity <- merge(nicheNet_P$ligandActivityTable,nicheNet_AD$ligandActivityTable,by="test_ligand",suffixes=c("_P","_AD"),all=TRUE)

segments <- c("CD8_AD","CD8_P","CD4_AD","CD4_P","KRT_P","KRT_AD")

expDat <- lapply(segments,function(x)
target_spatialData[ ,pData(target_spatialData)$Annotation %in% x,] %>%
  assayDataApply(2, FUN = log, base = 2, elt = "q_norm") %>% rowMeans(.) %>%
  stack %>%
  column_to_rownames("ind"))

expDat <- bind_cols(expDat)
colnames(expDat)<- segments

combinedActivity <- merge(combinedActivity,expDat,by.x="test_ligand",by.y="row.names")

combinedActivity <- combinedActivity[order(combinedActivity$aupr_corrected_P,decreasing = TRUE),]

best_upstream_ligands_P <- nicheNet_P$ligandActivityTable %>%
  top_n(10, aupr_corrected) %>% 
  arrange(-aupr_corrected) %>%
  pull(test_ligand)

best_upstream_ligands_AD <- nicheNet_AD$ligandActivityTable %>% 
  top_n(10, aupr_corrected) %>% 
  arrange(-aupr_corrected) %>% 
  pull(test_ligand)

ligandActivity <- combinedActivity[c("aupr_corrected_AD","aupr_corrected_P")]
rownames(ligandActivity) <- combinedActivity$test_ligand
colnames(ligandActivity) <- c("AD","P")

ranks <- apply(-ligandActivity,2,rank)

combinedActivity <- combinedActivity[ combinedActivity$test_ligand %in% c(best_upstream_ligands_P,best_upstream_ligands_AD),]
ranks <- ranks[ rownames(ranks) %in% c(best_upstream_ligands_P,best_upstream_ligands_AD),]
ligandActivity <- ligandActivity[ rownames(ligandActivity) %in% c(best_upstream_ligands_P,best_upstream_ligands_AD),]

col_fun <- circlize::colorRamp2(c(0, max(ligandActivity$P)), c("white", "red"))
cm <-  ColorMapping(col_fun = col_fun)


ligandExp <- combinedActivity[ ,sort(segments)]
rownames(ligandExp) <- combinedActivity$test_ligand

rowCM <- circlize::colorRamp2(c(0, max(ligandExp)), c("white", "blue"))

row_ha <- rowAnnotation(
  df = ligandExp,
  name = "expression",
  col = list(
    CD8_AD = rowCM,
    CD8_P = rowCM,
    CD4_AD = rowCM,
    CD4_P = rowCM,
    KRT_P = rowCM,
    KRT_AD = rowCM
  ),
  show_legend = c(TRUE, rep(FALSE, 5)),
  annotation_legend_param = list(CD4_AD = list(title = "log2 Exp")),
  annotation_name_gp = gpar(fontsize = 16)
)

p <-
  Heatmap(
    ligandActivity,
    name = "activity" ,
    show_column_names = TRUE,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    right_annotation = row_ha,
    col = col_fun,
    row_names_gp = grid::gpar(fontsize = 16),
    column_names_gp = grid::gpar(fontsize = 16),
    cell_fun = function(j, i, x, y, w, h, fill) {
      grid.text(ranks[i, j], x, y, gp = gpar(fontsize = 16,fontface="bold"))
    }
)


heatmap <-  grid.grabExpr(draw(p,padding= unit(c(7, 12, 15, 2), "mm")))


saveRDS(list(heatmap=heatmap,LR_AD=nicheNet_AD$ligandReceptor,LR_P=nicheNet_P$ligandReceptor),file="results/nicheNetPlots.RDS")

write_xlsx(list(Pso=nicheNet_P$ligandActivityTable,AD=nicheNet_AD$ligandActivityTable),path = "results/nicheNet.xlsx")


```

## Ligand receptor plots

```{r}

r1 <- nicheNet_AD$ligandReceptor +
  scale_fill_gradient(
    low = "white",
    high = "red",
    labels = function(x)
      sprintf("%.1f", x)
  ) +
  ylab("Prioritised ligands in AD") +
  xlab("Receptors") +
  theme(
    text = element_text(size = 16, color = "black"),
    axis.title =  element_text(size = 22),
    legend.text = element_text(size = 12),
    legend.key.width = unit(1, "cm"),
    axis.text = element_text(size = 14, color = "black")
  )


r2 <-
  nicheNet_P$ligandReceptor +  scale_fill_gradient(
    low = "white",
    high = "red",
    labels = function(x)
      sprintf("%.1f", x)
  ) + ylab("Prioritised ligands in Psorasis") + xlab("Receptors") + theme(
    text = element_text(size = 16, color = "black"),
    axis.title =  element_text(size = 22),
    legend.text = element_text(size = 12),
    legend.key.width = unit(1, "cm"),
    axis.text = element_text(size = 14, color = "black")
  )


layout <-
"AACC
 BBCC"


g <-  r1 + r2 + heatmap + plot_annotation(tag_levels = 'A')  + plot_layout(design = layout) &
  theme(plot.tag = element_text(face = 'bold', size = 22))
save_plot("figures/Fig4.png",plot = g,base_height = 12,base_width = 12,bg="white")


row1 <- cowplot::plot_grid(heatmap,ncol=1,labels = LETTERS[1],label_size = 25)
row2 <- cowplot::plot_grid(r1,r2,ncol=1,labels = LETTERS[2:3],label_size = 25)
g <- cowplot::plot_grid(row1,row2,ncol=2,byrow = FALSE,rel_widths = c(0.6,1))
save_plot("figures/Fig4.png",plot = g,base_height = 14,base_width = 12,bg="white")


```